---
title: "NYPD Shooting Incident Analysis"
author: "Kostia"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction

This analysis examines NYPD shooting incident data to identify patterns and trends in gun violence across New York City. The dataset contains historical records of shooting incidents, including temporal, geographic, and demographic information.


## Data Import and Initial Exploration
```{r setup, message=FALSE, warning=FALSE}
# Load required packages
library(tidyverse)
library(lubridate)
library(knitr)
library(pROC)

# Set global options
knitr::opts_chunk$set(echo = TRUE)

# Import data
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
shooting_data <- read_csv(url)
```
```{r summary}
# Data structure
glimpse(shooting_data)

# Summary statistics
summary(shooting_data)

# Check for missing values
missing_summary <- shooting_data %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  arrange(desc(missing_count))

kable(missing_summary, caption = "Missing Values by Variable")
```

## Data Cleaning Strategy

Based on the initial exploration, here's the cleaning approach:
- Convert OCCUR_DATE to proper date format, extract OCCUR_TIME components,create temporal features: year, month, day of week
- Standardize BORO formatting, remove latitude, longitude, and address variables due to high proportion of missing values
- Standardize categorical levels 
- Unify all missing data as NA for consistency
- Address data quality issues (e.g., age groups like "1020", "224", "940")
- Convert STATISTICAL_MURDER_FLAG to logical
```{r cleaning}
# Execute cleaning operations
shooting_clean <- shooting_data %>%
  # Remove geographic detail columns
  select(-Latitude, -Longitude, -Lon_Lat, 
         -LOCATION_DESC, -X_COORD_CD, -Y_COORD_CD) %>%
  # Temporal cleaning
  mutate(
    OCCUR_DATE = mdy(OCCUR_DATE),
    OCCUR_TIME = hms(OCCUR_TIME),
    occur_hour = hour(OCCUR_TIME),
    occur_year = year(OCCUR_DATE),
    occur_month = month(OCCUR_DATE, label = TRUE),
    occur_dow = wday(OCCUR_DATE, label = TRUE),
    time_category = case_when(
      occur_hour >= 6 & occur_hour < 12 ~ "Morning",
      occur_hour >= 12 & occur_hour < 18 ~ "Afternoon",
      occur_hour >= 18 & occur_hour < 24 ~ "Evening",
      TRUE ~ "Night"
    )
  ) %>%
  # Geographic cleaning
  mutate(BORO = str_to_title(BORO)) %>%
  # Demographic cleaning - unify all missing as NA
  mutate(
    across(c(PERP_AGE_GROUP, VIC_AGE_GROUP), 
           ~case_when(
             . %in% c("1020", "224", "940", "1028") ~ NA_character_,
             . == "UNKNOWN" ~ NA_character_,
             TRUE ~ .
           )),
    across(c(PERP_SEX, VIC_SEX, PERP_RACE, VIC_RACE),
           ~if_else(. == "UNKNOWN", NA_character_, .))
  ) %>%
  # Convert murder flag
  mutate(STATISTICAL_MURDER_FLAG = as.logical(STATISTICAL_MURDER_FLAG)) %>%
  # Remove future dates
  filter(OCCUR_DATE <= Sys.Date())

# Verify cleaning results
cat("Original rows:", nrow(shooting_data), "\n")
cat("Cleaned rows:", nrow(shooting_clean), "\n")
```

## Analysis

The analysis focuses on identifying spatial and temporal patterns in shooting incidents across New York City. 

**Strategy**: Start with geographic distribution to understand volume disparities, then examine temporal patterns  to identify when incidents occur.

### Geographic Distribution
```{r viz1_borough}
# Shooting incidents by borough
shooting_clean %>%
  count(BORO) %>%
  ggplot(aes(x = reorder(BORO, n), y = n, fill = BORO)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Shooting Incidents by Borough",
       x = NULL,
       y = "Number of Incidents") +
  theme_minimal() +
  theme(text = element_text(size = 12))
```

### Temporal Trends - City Wide
```{r viz2_temporal_citywide}
# Overall temporal trend
shooting_clean %>%
  count(occur_year, occur_month) %>%
  mutate(date = make_date(occur_year, occur_month)) %>%
  ggplot(aes(x = date, y = n)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_smooth(method = "loess", se = TRUE, color = "darkred", linewidth = 0.8) +
  labs(title = "Shooting Incidents Over Time - New York City",
       x = "Date",
       y = "Number of Incidents") +
  theme_minimal() +
  theme(text = element_text(size = 12))
```

### Temporal Trends - By Borough
```{r viz3_temporal_borough}
# Temporal trend with all boroughs overlaid
shooting_clean %>%
  count(BORO, occur_year, occur_month) %>%
  mutate(date = make_date(occur_year, occur_month)) %>%
  ggplot(aes(x = date, y = n, color = BORO)) +
  geom_line(linewidth = 0.9) +
  labs(title = "Shooting Incidents Over Time by Borough",
       x = "Date",
       y = "Number of Incidents",
       color = "Borough") +
  theme_minimal() +
  theme(text = element_text(size = 12),
        legend.position = "bottom")
```

### Time of Day Patterns - City Wide
```{r viz4_timeofday_citywide}
# Overall time of day heatmap
shooting_clean %>%
  count(occur_dow, occur_hour) %>%
  ggplot(aes(x = occur_hour, y = occur_dow, fill = n)) +
  geom_tile() +
  scale_fill_gradient(low = "lightyellow", high = "darkred") +
  labs(title = "Shooting Incidents by Day of Week and Hour - New York City",
       x = "Hour of Day",
       y = "Day of Week",
       fill = "Incidents") +
  theme_minimal() +
  theme(text = element_text(size = 12))
```

### Time of Day Patterns - By Borough
```{r viz5_timeofday_borough}
# Time of day heatmap by borough
shooting_clean %>%
  count(BORO, occur_dow, occur_hour) %>%
  ggplot(aes(x = occur_hour, y = occur_dow, fill = n)) +
  geom_tile() +
  scale_fill_gradient(low = "lightyellow", high = "darkred") +
  facet_wrap(~BORO, ncol = 3) +
  labs(title = "Shooting Incidents by Day of Week and Hour - By Borough",
       x = "Hour of Day",
       y = "Day of Week",
       fill = "Incidents") +
  theme_minimal() +
  theme(text = element_text(size = 10),
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8))
```

### Incident Lethality - City Wide
```{r viz6_lethality_citywide}
# Proportion of fatal shootings over time - citywide
shooting_clean %>%
  count(occur_year, STATISTICAL_MURDER_FLAG) %>%
  group_by(occur_year) %>%
  mutate(total = sum(n),
         fatal = sum(n[STATISTICAL_MURDER_FLAG == TRUE])) %>%
  filter(STATISTICAL_MURDER_FLAG == TRUE) %>%
  mutate(fatal_proportion = fatal / total) %>%
  ggplot(aes(x = occur_year, y = fatal_proportion)) +
  geom_line(color = "darkred", linewidth = 1.2) +
  geom_point(color = "darkred", size = 2) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Fatal Shootings Over Time - New York City",
       x = "Year",
       y = "Fatal Shooting Rate") +
  theme_minimal() +
  theme(text = element_text(size = 12))
```

### Incident Lethality - By Borough
```{r viz7_lethality_borough}
# Proportion of fatal shootings over time - by borough
shooting_clean %>%
  count(BORO, occur_year, STATISTICAL_MURDER_FLAG) %>%
  group_by(BORO, occur_year) %>%
  mutate(total = sum(n),
         fatal = sum(n[STATISTICAL_MURDER_FLAG == TRUE])) %>%
  filter(STATISTICAL_MURDER_FLAG == TRUE) %>%
  mutate(fatal_proportion = fatal / total) %>%
  ggplot(aes(x = occur_year, y = fatal_proportion, color = BORO)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Fatal Shootings Over Time by Borough",
       x = "Year",
       y = "Fatal Shooting Rate",
       color = "Borough") +
  theme_minimal() +
  theme(text = element_text(size = 12),
        legend.position = "bottom")
```
### Analysis Summary

The data reveals that Brooklyn and the Bronx consistently show the highest volumes, with a notable spike across all boroughs around 2020-2021 before declining. Shootings concentrate heavily in late evening and overnight hours on weekends. The fatal shooting rate has remained relatively stable between 18-25% across most years,but there is a strong variation between boroughs; some showing sustained higher lethality rates than others. Thus, it looks like temporal and geographic factors play significant roles in shooting incidents.




## Modeling

I will build a logistic regression model to predict shooting lethality (fatal vs non-fatal outcomes). Based on the visualizations, we expect borough, time of day, and temporal factors to influence lethality rates. Even though excluded form visualisations, victim age group will also be included in the model.

Since fatal shootings represent less than 20% of incidents, we'll use a 0.2 probability threshold for classification rather than the standard 0.5. This way I will try to account for the class imbalance and, hopefully, produces more realistic predictions.

### Model Development
```{r model_prep}
# Prepare modeling dataset - remove NAs in key variables
model_data <- shooting_clean %>%
  filter(!is.na(VIC_AGE_GROUP)) %>%
  mutate(
    # Relevel factors for interpretation
    BORO = relevel(factor(BORO), ref = "Manhattan"),
    time_category = relevel(factor(time_category), ref = "Morning"),
    VIC_AGE_GROUP = factor(VIC_AGE_GROUP)
  )

# Check sample size
cat("Modeling dataset:", nrow(model_data), "observations\n")
cat("Fatal incidents:", sum(model_data$STATISTICAL_MURDER_FLAG), "\n")
cat("Non-fatal incidents:", sum(!model_data$STATISTICAL_MURDER_FLAG), "\n")
cat("Baseline fatal rate:", round(mean(model_data$STATISTICAL_MURDER_FLAG), 3), "\n")
```
```{r model_fit}
# Fit logistic regression
lethality_model <- glm(
  STATISTICAL_MURDER_FLAG ~ BORO + time_category + VIC_AGE_GROUP + occur_year,
  data = model_data,
  family = binomial(link = "logit")
)

# Model summary
summary(lethality_model)
```

### Model Results
```{r model_coefficients}
# Extract and format coefficients
coef_table <- broom::tidy(lethality_model, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  arrange(desc(abs(estimate - 1)))

kable(coef_table, 
      caption = "Logistic Regression Coefficients (Odds Ratios)",
      col.names = c("Predictor", "Odds Ratio", "Std Error", "Z-value", "P-value", "CI Low", "CI High"))
```
```{r model_visualization}
# Visualize key odds ratios
coef_table %>%
  filter(term != "(Intercept)") %>%
  filter(p.value < 0.05) %>%
  ggplot(aes(x = reorder(term, estimate), y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Significant Predictors of Shooting Lethality (Odds Ratios)",
       subtitle = "Values > 1 indicate increased odds of fatal outcome",
       x = NULL,
       y = "Odds Ratio (95% CI)") +
  theme_minimal() +
  theme(text = element_text(size = 11))
```

### Model Evaluation
```{r model_evaluation}
# Predictions and performance metrics
model_data <- model_data %>%
  mutate(
    predicted_prob = predict(lethality_model, type = "response"),
    predicted_class = if_else(predicted_prob > 0.2, TRUE, FALSE)
  )

# Confusion matrix
confusion <- table(Predicted = model_data$predicted_class, 
                   Actual = model_data$STATISTICAL_MURDER_FLAG)

cat("Confusion Matrix:\n")
print(confusion)
cat("\n")

# Calculate metrics safely
accuracy <- sum(diag(confusion)) / sum(confusion)

# Check if we have both predicted classes
if (nrow(confusion) == 2 && ncol(confusion) == 2) {
  sensitivity <- confusion[2,2] / sum(confusion[,2])
  specificity <- confusion[1,1] / sum(confusion[,1])
  precision <- confusion[2,2] / sum(confusion[2,])
} else {
  # If model only predicts one class
  sensitivity <- ifelse(nrow(confusion) == 2, confusion[2,2] / sum(confusion[,2]), 0)
  specificity <- ifelse(nrow(confusion) == 2, confusion[1,1] / sum(confusion[,1]), 1)
  precision <- ifelse(nrow(confusion) == 2, confusion[2,2] / sum(confusion[2,]), 0)
}

metrics_df <- data.frame(
  Metric = c("Accuracy", "Sensitivity (Recall)", "Specificity", "Precision"),
  Value = round(c(accuracy, sensitivity, specificity, precision), 3)
)

kable(metrics_df, caption = "Model Performance Metrics")
```
```{r roc_curve}
# ROC curve
library(pROC)
roc_obj <- roc(model_data$STATISTICAL_MURDER_FLAG, model_data$predicted_prob)

cat("AUC:", round(auc(roc_obj), 3), "\n")

# Plot ROC curve
plot(roc_obj, 
     main = "ROC Curve for Lethality Prediction Model",
     col = "darkred", 
     lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "gray")
text(0.6, 0.2, paste("AUC =", round(auc(roc_obj), 3)), cex = 1.2)
```

### Model Interpretation

The model identifies borough, time of day, victim age, and year as factors associated with shooting lethality, but the predictive power is low. The results show weak effects and the model barely outperforms baseline predictions, indicating more sophisticated modeling approaches (interaction terms, additional predictors like precinct-level data, or machine learning methods) are needed to meaningfully predict fatal outcomes.

## Bias Identification

**Data Collection Bias**
- Not all shootings are reported to police. Incidents in certain neighborhoods or involving certain demographics may be systematically underreported
- Victim cooperation varies by community, affecting incident classification
- Police presence and reporting practices differ across precincts, thus, skewing geographic patterns

**Missing Data Patterns**
- Perpetrator demographic information is heavily missing (age, race, sex), limiting analysis of offender characteristics
- Missing data may not be random but show systematic racism and other social factors.
- Geographic coordinates are incomplete, preventing spatial analysis at more granular level 

**Definitional and Classification Issues**
- STATISTICAL_MURDER_FLAG reflect suspitions and not convictions
